---
title: "Loading NHIS Data"
author: "Kevin R Foster"
output: html_document
---

This uses data from CDC's [National Health Interview Survey](http://www.cdc.gov/nchs/nhis/quest_data_related_1997_forward.htm).

This Rmd file is an abbreviated version of one for ATUS that I wrote.  There are other files in the zip: personx.dat is the original data from CDC, personx.sps.txt is their file to upload to SPSS and has coding notes and variable labels, then there are a few minor files that I created to help load.

Step 1 is to input the data.


```{r}
dat_nhis <- read.csv("personsx.csv")
summary(dat_nhis)
```


There are 2 steps from the .sps program: 1. associating the short name with a longer name, and 2. giving the codes for what the data means.

For example it is easy to figure out that in "SRVY_YR" which is Survey Year the value 2003 indicates the year 2003. Duh. But in "REGION" what do the numbers 1 to 4 mean? The .sps program tells that
 | 1 | "Northeast"
 | 2 | "Midwest"
 | 3 | "South"
 | 4 | "West"
and similarly for other variables.

Must be really careful with false friends like "AGE" where it goes in what seems like a perfectly legible way, so the code "35" means that the person is age 35 etc. But at the end 85 is the value for top-coding.
```{r}

# lazy recoding
attach(dat_nhis)

REGION <- as.factor(REGION)
levels(REGION) <- c("Northeast","Midwest","South","West")
female <- as.numeric(SEX == 2)
Hispanic <- as.numeric(ORIGIN_I == 1)
Hispan_PR <- as.numeric(HISPAN_I == "01")
Hispan_Mex <- as.numeric(HISPAN_I == "02") + as.numeric(HISPAN_I == "03")
Hispan_DR <- as.numeric(HISPAN_I == "05")
AfAm <- as.numeric(RACRECI3 == 2)
Asian <- as.numeric(RACRECI3 == 3)
RaceOther <- as.numeric(RACRECI3 == 4)
# RRP is Reference Person - 1 for ref, etc
AGE_P <- as.numeric(AGE_P)
married <- as.numeric((R_MARITL == 1) | (R_MARITL == 2) | (R_MARITL == 3))
widowed <- as.numeric((R_MARITL == 4))
divorc_sep <- as.numeric((R_MARITL == 5) | (R_MARITL == 6))
disabl_limit <- as.numeric(LA1AR == 1)
person_healthstatus <- as.factor(PHSTAT)
levels(person_healthstatus) <- c("Excellent","Very good", "Good", "Fair", "Poor", "Refused", "Not ascertained", "Don't know")
# HOSPNO <- as.numeric(HOSPNO)
# HOSPNO[HOSPNO > 996] <- NA
NOTCOV <- as.numeric(NOTCOV == 1)
MEDICARE <- as.numeric((MEDICARE == 1) | (MEDICARE == 2))
MEDICAID <- as.numeric((MEDICAID == 1) | (MEDICAID == 2))
private_ins <- as.numeric((PRIVATE == 1) | (PRIVATE == 2))
SCHIP <-  as.numeric((SCHIP == 1) | (SCHIP == 2))
sptn_medical <- as.factor(HCSPFYR) # amount spent on medical
levels(sptn_medical) <- c("zero","under 500","500-1999","2000-2999","3000-4999","5000+","refused",NA,NA)
borninUSA <- as.numeric(PLBORN == 1)
region_born <- as.factor(REGIONBR)
levels(region_born) <- c("US","Mex Cent Am Caribb","S Am","Eur","former USSR","Africa","MidE","India subc","Asia","SE Asia","Elsewhere","unknown")
educ_nohs <- as.numeric(EDUC1 < 12)
educ_hs <- as.numeric((EDUC1 == 13) | (EDUC1 == 14))
educ_smcoll <- as.numeric(EDUC1 == 15)
educ_as <- as.numeric((EDUC1 > 15) & (EDUC1 < 18))
educ_bach <- as.numeric(EDUC1 == 18)
educ_adv <- as.numeric((EDUC1 > 18) & (EDUC1 < 96))
veteran_stat <- as.numeric(ARMFEV == 1)
inworkforce <- as.numeric(DOINGLWP < 5)
WRKHRS2 <- as.numeric(WRKHRS2)
earn_lastyr <- as.factor(ERNYR_P)
levels(earn_lastyr) <- c("0","$01-$4999","$5000-$9999","$10000-$14999","$15000-$19999","$20000-$24999","$25000-$34999","$35000-$44999","$45000-$54999","$55000-$64999","$65000-$74999","$75000 and over",NA,NA,NA)

data_use1 <- data.frame(NOTCOV, educ_nohs, educ_hs, educ_smcoll, educ_as, educ_bach, educ_adv, AGE_P, female, AfAm, Asian, RaceOther,Hispanic, Hispan_PR, Hispan_Mex, Hispan_DR, married, widowed, divorc_sep, REGION, borninUSA, region_born,veteran_stat, inworkforce, ERNYR_P, disabl_limit, person_healthstatus, MEDICARE, MEDICAID, private_ins, RRP, HHX, FMX, FPX, SCHIP, sptn_medical)

detach()

```
Note that I'm leaving much of the data and not using it.  If someone wanted to go back and add a variable to the data_use1 data frame, it would be easy enough.

```{r}
save(data_use1, file = "NHIS_2017.RData")
```

Later instead of running all of those lines you can just read in the data file with just this code,
```{r}
load("NHIS_2017.RData")
```


